---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
activity <- read.csv(unz("activity.zip", "activity.csv"), colClasses=c("numeric","Date","numeric"))

# aggregates data into number of steps per day
stepsPerDay <- aggregate(steps ~ date, data=activity, sum)
# Averages 5 min intervals over all days for the second plot
averageActivity <- aggregate(steps ~ interval, data=activity, mean)
```

## What is mean total number of steps taken per day?
```{r}
steps <- stepsPerDay$steps
hist(steps, breaks=20)
```

## The mean of steps per day is
```{r}
mean(steps, na.rm=TRUE)
```

## The median of steps per day is
```{r}
median(steps, na.rm=TRUE)
```

## What is the average daily activity pattern?
```{r}
with(averageActivity, plot(interval, steps, type="l"))
```

## The 5-minute interval that contains the maximum number of steps over a full day is:
```{r}
averageActivity[which.max(averageActivity$steps),"interval"]
```


## Imputing missing values
For imputing I simply use the mean of steps for each interval across all days in the dataset.

## The number of rows that have missing values is
```{r}
# get the rows that have missing values
missing <- activity[!complete.cases(activity),]
# print number of rows with missing values
dim(missing)[1]
```

## Creating the dataset `imputedActivity` with imputed values
```{r}
imputable <- activity[!complete.cases(activity),]

# Notice the usage of averageActivity data that was computed earlier
imputedSteps <- merge(imputable, averageActivity, by.x="interval", by.y="interval")
imputedSteps$steps <- imputedSteps$steps.y
imputedSteps$steps.x <- NULL
imputedSteps$steps.y <- NULL

imputedActivity <- activity
imputedActivity[!complete.cases(imputedActivity),"steps"] <- imputedSteps$steps
```




## Histogram of total steps taken each day after imputation
```{r}

steps <- aggregate(steps ~ date, data=imputedActivity, sum)$steps
hist(steps, breaks=20)
```

## The imputed mean of steps per day is
```{r}
mean(steps, na.rm=TRUE)
```

## The imputed median of steps per day is
```{r}
median(steps, na.rm=TRUE)
```


The impact of imputing is that the mean basically stays the same but median grows by about 250.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
# Create a factor variable weekday that indicates whether the date is a weekday or weekend
imputedActivity$weekday = ifelse(weekdays(imputedActivity$date) == "Saturday" |
                                         weekdays(imputedActivity$date) == "Sunday",
                                 "weekend", "weekday")

# Calculate avg steps per interval using factor weekday
library(plyr)
stepsPerWeekday <- ddply(imputedActivity, c("interval","weekday"),summarize,
                         steps=mean(steps,na.rm = TRUE))

library(lattice)
xyplot(steps~interval|weekday,data=stepsPerWeekday, layout=c(1,2), type="l")
```

